SubGenome Analysis

### 2017-07-27 R/SubGenome Analysis.Rmd
Basic import done, ready to start analysis portion

---
title: "SubGenome Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
Some preprocessing was needed to generate the input files for this project.  We begin with SynMap output based on the Schnable 2011 paper (see https://www.evernote.com/shard/s303/nl/52987123/1ce74f11-98a0-4854-b363-3172e79eeae3 for details).

This resulted in the file (31607_34889.CDS-CDS.lastz.tdd10.cs0.filtered.dag.all.go_D20_g10_A5.aligncoords.gcoords.ks).

We parse this file to a flat format using a perl script (parseKsKnFile.pl)

(Do this in command line until I figure out how to run it from R)

> perl parseKsKnFile.pl ../Data/31607_34889.CDS-CDS.lastz.tdd10.cs0.filtered.dag.all.go_D20_g10_A5.aligncoords.gcoords.ks > ../Data/sorghum\ vs\ maize.tab

```{r parseSynMap, include=FALSE}
cmd <- paste("perl", "parseKsKnFile.pl", "31607_34889.CDS-CDS.lastz.tdd10.cs0.filtered.dag.all.go_D20_g10_A5.aligncoords.gcoords.ks")
system(cmd)
```

The data is now prepared for import and cleaning.  We will import the file and remove any incomplete lines.

```{r importSynMapData, echo=FALSE}
library(readr)
library(plyr)
library(reshape2)
library(ggplot2)

## Import data and remove incomplete rows
syntelogs.import <- read_delim("../Data/sorghum vs maize.tab", "\t", escape_double = FALSE, trim_ws = TRUE)
syntelogs.complete <- syntelogs.import[complete.cases(syntelogs.import),]

sprintf("We have imported %d lines from the parsed SynMap file and removed %d incomplete rows resulting in %d syntenic genes.", nrow(syntelogs.import), nrow(syntelogs.import)- nrow(syntelogs.complete), nrow(syntelogs.complete))
```

We need to select a cutoff ks value to differentiate between the most recent alpha duplication event and the previous beta duplication.

```{r pickKSCutoff, echo=FALSE}
## Pick a ks value which separates alpha duplication from beta duplication (between the first two peaks)
log10_ks_cutoff <- -0.575
qplot(log10(syntelogs.complete$ks), geom="histogram", binwidth=.01)  + geom_vline(mapping = NULL, data = NULL, xintercept = log10_ks_cutoff, na.rm = FALSE, show.legend = NA)
ks_cutoff <- 10^log10_ks_cutoff
#ks_cutoff <- 0.4234 #include most of the first 2 peaks
```

@@

(but how to group subgenomes over multiple chromosomes?)

@@

```{r writeSubgenomesToFile, echo=FALSE}
## Now recover which genes are in these sets
# subset(syntelogs.orthologs, count>=12 & median<=ks_cutoff & org_chr1=="a31607_1" & org_chr2=="b34889_1")
# orthologs.blocks_groups <- syntelogs.orthologs[syntelogs.orthologs$count>=12 & syntelogs.orthologs$median<=ks_cutoff,]
subgenome.groups <- rbind(merge(syntelogs.orthologs, orthologs.sub1, all.x = F), merge(syntelogs.orthologs, orthologs.sub2, all.x = F))
subgenome <- merge(syntelogs.complete, subgenome.groups, all.x = F)
write.table(subgenome, "../Output/recoveredSubGenomes.tab", sep="\t")
```

>>>>

---
title: "SubGenome Analysis"
output: html_document
params:
  fileParser: "../Perl/parseKsKnFile.pl"

  #v1 vs. v1
  synMapOutputFile: "6807_8082.CDS-CDS.lastz.tdd10.cs0.filtered.dag.all.go_D20_g10_A5.aligncoords.gcoords.ks"
  inputDataFile: "../Data/sorghum_v1_vs_maize_v1.tab"
  outputFile: "../Output/recoveredSubGenomes_sorghum_v1_vs_maize_v1.tab"
  log10_ks_cutoff: -0.575

  #v3.1 vs v4+rejected
  #synMapOutputFile: ""
  #outputFile: "../Output/recoveredSubGenomes_sorghum_v3.1_vs_maize_v4+rejected.tab"
  #inputDataFile: "../Data/sorghum_v3.1_vs_maize_v4+rejected.tab"
  #log10_ks_cutoff: -0.575
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(readr)
library(plyr)
library(reshape2)
library(ggplot2)
library(readxl)
```
```{r preamble, echo=FALSE}
paste("This analysis is using the file", params$inputDataFile)
```

# Using Sorghum as an outgroup to find syntenic regions in Maize: regenerating subgenome assignments in maize
We begin with SynMap output based on the [Schnable 2011](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3053962/) paper (see [my notes](https://www.evernote.com/shard/s303/nl/52987123/1ce74f11-98a0-4854-b363-3172e79eeae3) for details on parameter and download options).

## Preprocessing to generate a tidy dataset
We parse the SynMap output file with ks/kn values (i.e. a DAGChainer file) to a flat format using a perl script (parseKsKnFile.pl) and import the data into R.  Some ks and kn values are empty, so we clean the dataset by removing all incomplete observations.

> manually perform the parse, correct the header lines, and run the import lines until pipeline is debugged
> perl parseKsKnFile.pl ../Data/6807_8082.CDS-CDS.lastz.tdd10.cs0.filtered.dag.all.go_D20_g10_A5.aligncoords.gcoords.ks > ../Data/sorghum_v1_vs_maize_v1.tab

```{r importSynMapData, include=FALSE}
cmd <- paste("perl", params$fileParser, params$synMapOutputFile)
system(cmd)
syntelogs.import <- read_delim(params$inputDataFile, "\t", escape_double = FALSE, trim_ws = TRUE)
syntelogs.complete <- syntelogs.import[complete.cases(syntelogs.import),]
```
```{r reportSynMapData, echo=FALSE}
# Report results of import
sprintf("We have imported %d lines from the parsed SynMap file and removed %d incomplete rows resulting in %d syntenic genes.", nrow(syntelogs.import), nrow(syntelogs.import)- nrow(syntelogs.complete), nrow(syntelogs.complete))
```

We need to select a cutoff value for ks (synonymous mutation rate) to differentiate between the most recent alpha duplication event and the previous beta duplication. If the vertical line separates the first two peaks sufficently, it is probably a good enough value.  If not, we need to pick a new cutoff value for ks.

```{r pickKSCutoff, echo=FALSE}
ks_cutoff <- 10^params$log10_ks_cutoff
qplot(log10(syntelogs.complete$ks), geom="histogram", binwidth=.01)  + geom_vline(mapping = NULL, data = NULL, xintercept = params$log10_ks_cutoff, na.rm = FALSE, show.legend = NA)
```

@@

> But how to group subgenomes over multiple chromosomes?

@@

```{r writeSubgenomesToFile, echo=FALSE}
## Now recover which genes are in these sets
subgenome.groups <- rbind(merge(syntelogs.orthologs, orthologs.sub1, all.x = F), merge(syntelogs.orthologs, orthologs.sub2, all.x = F))
subgenome <- merge(syntelogs.complete, subgenome.groups, all.x = F)
write.table(subgenome, params$outputFile, sep="\t")
```

We also want to know how we did compared to the paper itself, to see if this method is faithfully reproducing their results.
```{r compareResultsToPaper, echo=FALSE}
sd01 <- read_excel("../Data/sd01.xls")
```
